\chapter{Introduction}\label{C:intro}
This work is concerned with the analysis and implementation of tools to support the automated refactoring of Rust programs. The resulting refactoring tool built is written in Rust and [available from Github under an MIT license]. The purpose of this chapter is to describe the overall work presented in this document and to begin to describe aspects of the motivation for pursuing this work. In Section \ref{S:rustintro} and Section \ref{S:refactorintro} the context of the project is introduced with Rust and refactoring. In Section \ref{S:contrib}, the contributions made by this work are described. In Section \ref{S:implemented}, the definite list of refactorings supported by the tool is described, along with a brief description of each. Lastly, in Section \ref{S:outline}, a summary of the subsequent chapters and the remaining document is given.

%In Chapter \ref{C:us} we explain how to use this document, and the \texttt{vuwproject} style. In Chapter \ref{C:ex} we say some things about \LaTeX, and in Chapter \ref{C:con} we give our conclusions.

\section{Introducing the Rust programming language}\label{S:rustintro}

% http://blog.rust-lang.org/2015/05/15/Rust-1.0.html

The Rust programming language, as of May 2015, reached the 1.0 milestone.  As a modern memory-safe systems programming language, the aim for the language is to provide reliable and efficient systems by combining the performance of low level control, with the convenience and guarantees of higher level constructs. All of this is achieved without a garbage collector or runtime and allows interoperability with no overhead with C. Rust enforces an ownership system to restrict the duplication of references through borrowing and lifetimes, while still aiming for `zero cost abstractions'. Using this techniques, Rust prevents dangling pointers and whole class of related issues concerning iterator invalidation, concurrency and more [ref https://doc.rust-lang.org/book/README.html].

Rust has been a number of years in the making [ref date?], but only with this public 1.0 release has many of the standard library API stabilized. In order to further present itself as a stable and mature language, there is still another component which has yet to be adequately tackled: tooling. Developers commonly use a rich set of tools to simplify the task of building complex software with IDE, editors and debuggers. Such tools can improve productivity and providing tools specific to a language may help to encourage adoption.

\section{The issue of refactoring}\label{S:refactorintro}
[Should this summarize the next chapter or not?] Refactoring is the act of performing functionality preserving code transformation. Traditionally, these transformations needed to be performed manually, but in recent years, a number of tools to aid and automate refactorings have arisen in many programming languages such as Java or C++ [cite some source?]. Manual transformations, including editor search-and-replace are potentially prone to error and so performing tool-assisted refactoring which guarantees some measure of correctness often provides much greater confidence in changes.

\section{Contributions of this work}\label{S:contrib}
The main purpose of this work is to produce a proof-of-concept refactoring tool which utilizes existing infrastructure made available by the Rust compiler. With continuing expectations for change within the language and compiler and a number of API are still undergoing stabilization, a refactoring tool would prove beneficial for users at risk of changes breaking their code underneath them. In particular, API changes which convert global functions e.g. foo(A, B) to A.foo(B), to methods acting on an object are common and providing hopefully this work provides some insight into the necessary work to achieve such a refactoring. 

Exploring the necessary barriers to implementing a refactoring tool, proposals or implementation of extensions to the Rust compiler are brought forth and help to identify current shortcomings with providing refactoring tools for the Rust programming language and work necessary to improve the capability of Rust to support refactoring. Additionally, the current extent of the ability for the compiler to facilitate refactoring is evaluated and a number of internal or unstable API are assessed. The compile driver API which provides the ability for Rust code to invoke the compiler is a primary example of code which has seen little real world usage. The name resolution module within the compiler is another key example used in this work for performing refactoring, being used in a manner not originally intended or imagined despite providing part of the functionality required for implementing such a tool.

This work presents an actual refactoring tool which may be run to provide a set of refactorings which may continue to be extended in the future. The list of refactorings supported are listed in the next section and described in detail in Chapter \ref{C:wd}. This work also attempts to provide initial evaluations as to the effectiveness and practicality of the current tool. Furthermore it aims to identify the most useful refactorings not currently supported by the tool and the most useful future changes required within the compiler for further iterations on the current work.

\section{Implemented refactorings}\label{S:implemented}
[Insert brief summary here]

\begin{itemize}
\item \bf{Renaming local variables} --
\item \bf{Renaming global variables} --
\item \bf{Renaming functions} --
\item \bf{Renaming methods} --
\item \bf{Renaming structs} -- 
\item \bf{Renaming enumerations} --
\end{itemize}

\section{Document outline}\label{S:outline}
[Should this be bullet pointed - should it be more descriptive?]
In Chapter \ref{C:wd} we explain the overall Rust refactoring tool, the major design decisions made [while in Chapter ?]/[and] we discuss the specifics for the implementation with a more indepth focus on Rust behaviour. In Chapter \ref{C:eval}, we discuss opportunities for evaluation of the tool and the current extent of testing for refactorings provided by the tool and overall correctness. In Chapter \ref{C:future}, we propose a number of further extensions to the tool, to the Rust compiler and in general, outline areas of future work while providing useful insights gained during the course of this work. Lastly, Chapter \ref{C:con} presents a summary, and gives conclusions about this work.